Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-06 _tex/index.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. The present study primarily focuses on two: the
over-reliance on the assumptions of Thurstone's Case V in the
statistical analysis of CJ data, and the apparent disconnect between
CJ's trait measurement and hypothesis testing. The following sections
begin with a brief overview of Thurstone's theory and a detailed
discussion of these issues. Subsequently, the study introduces a
theoretical model for CJ that builds upon Thurstone's theory, alongside
its statistical translation, designed to address the two concerns
simultaneously.

@@

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait. According to the theory, the discriminal process
for each stimulus follows a Normal distribution, where its mode (mean),
referred to as the \emph{modal discriminal process}, indicates the
stimulus' position on the trait continuum, and its dispersion, known as
the \emph{discriminal dispersion}, reflects the variability of the
stimulus' perceived trait.

For instance, Figure~\ref{fig-discriminal_process} illustrates the
discriminal process distributions along a quality trait continuum for
two written texts. The figure shows that these processes follow a Normal
distribution. Moreover, it depicts differences in the texts' positions
along the quality trait continuum, where text B is positioned further
along the continuum than text A, as indicated by their modal discriminal
processes (\(S_{B}\) and \(S_{A}\)). Finally, it highlights differences
in the texts' discriminal dispersions (\(\sigma_{B}\) and
\(\sigma_{A}\)), showing that text B exhibits a greater variability in
its perceived quality than text A, as reflected by its wider
distribution.

@@

\caption{\label{fig-discriminal_process}Example distributions of the
discriminal processes for two written texts}

@@

However, because the discriminal process of a single stimulus is not
directly observable, the theory introduces the \emph{law of comparative
judgment}. This law posits that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This emphasizes that the outcome of
a pairwise comparison likely depends on the relative distance between
stimuli rather than their absolute positions on the trait continuum.

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution
\citep{Andrich_1978}. The mode (mean) of this distribution, representing
the (average) relative separation, is given by the difference between
the modal discriminal processes of the stimuli
\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
distribution, reflecting the variability in the relative separation, is
calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) denote the discriminal
dispersions of the stimuli, while \(\rho\) represents the correlation
between their discriminal processes. This correlation quantifies the
dependence of the judge's perception of the trait in one stimulus on his
perception of the same trait in another.

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}, assuming a correlation of
\(\rho = 0.6\). The figure reveals that, under these conditions, the
judge perceives text B as having significantly higher quality than text
A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
a result, the dichotomous outcome of this comparison would likely favor
text B over text A.

@@

\caption{\label{fig-discriminal_difference}Example distribution of the
discriminal difference for the two texts shown in
Figure~\ref{fig-discriminal_process}, assuming a correlation of 0.6}

@@

Notably, the correlation between the discriminal processes, \(\rho\),
plays a pivotal role in determining the comparison outcome by shaping
the distribution of the discriminal difference between the stimuli.
Specifically, as the correlation increases, reflecting a stronger
dependence of the judge's perception of quality in one stimulus on his
perception of the other, the distribution of the discriminal difference
narrows. This narrowing ultimately impacts the area under the curve that
determines the comparison outcome and, consequently, the conclusions
drawn from this outcome.

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the texts
depicted in Figure~\ref{fig-discriminal_process}. Since the texts differ
in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of the area under the curve,
\(P(B > A)\), that lies above zero compared to curves with lower
correlations. Moreover, although the figure does not illustrate this
scenario, it is reasonable to infer that if the texts had similar or
identical quality levels, higher correlations would likely reduce the
chance of the discriminal difference distinctly favoring one text over
the other. This probability reduction occurs because the distribution of
the discriminal difference would become more narrowly centered around
zero.

\caption{\label{fig-correlation}The effect of correlation on the
distribution of the discriminal difference of the same two written text}

\section{Three critical issues in CJ
literature}\label{sec-theory-issues}

@@

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
be suitable for applications beyond the psycho-physical contexts for
which it was created. Thurstone himself cautioned that its use ``should
not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
could prove problematic in the presence of complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}. Consequently,
given that modern CJ applications frequently involve these traits and
stimuli, two main assumptions of Case V and, by extension, of the BTL
model may not consistently hold in theory or practice: the equal
dispersion and zero correlation between stimuli.

On the one hand, the assumption of \emph{equal dispersion between
stimuli} suggests that the variability of the perceived trait remains
consistent across all stimuli. However, Thurstone contended that this
assumption may not hold when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
features. Indeed, evidence of such violation may already exist in the CJ
literature as misfit statistics. These statistics measure the judgment
discrepancies associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than
comparisons involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that the discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
their discriminal processes also exhibit more variation than other
texts. Notably, this reasoning also applies to ``misfit'' judges, whose
evaluations reflect substantial deviations from the shared consensus due
to the unique characteristics of the stimuli or the judges themselves.

Nonetheless, assuming equal dispersions between stimuli despite its
violation can lead Case V (and the BTL model) to overlook critical
differences in the reliability of the trait across stimuli, resulting in
inaccurate conclusions about the trait's estimates
\citep[pp.~370]{McElreath_2020}. Furthermore, if researchers acknowledge
that misfit statistics help identify these critical differences, the
common practice in the CJ literature of excluding stimuli based on these
statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are unpredictable, as they
depend on the specific stimuli researchers exclude from the analysis.
Taken together, these oversights undermine the reliability of the trait
and ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.

Fortunately, the statistical literature offers a solution to address the
abovementioned issues. This solution involves using models that extend
traditional approaches to account for the different variability in the
stimuli, referred to as over-dispersed models
\citep[chap.~12]{McElreath_2020}.

On the other hand, the assumption of \emph{zero correlation between
stimuli} implies that, during a pairwise comparison, a judge's
perception of quality in one text does not influence his perception of
the same trait in another text (see Section~\ref{sec-thurstone_theory}).
Thurstone attributed this independence to the cancellation of potential
judges' biases, driven by two opposing and equally weighted effects
occurring during the pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
demonstrated this cancellation using the BTL model under the assumption
of discriminal processes with additive biases. However, it is easy to
imagine at least two scenarios where the zero correlation assumption
almost certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

@@

In any case, the psychometric and statistical literature emphasizes the
importance of addressing the factors that create dependencies between
stimuli, as failing to do so can lead to inaccurate conclusions about a
trait's reliability and, by extension, its validity. For instance,
neglecting additional traits relevant to the stimuli, such as judges'
biases, often leads to a dimensional mismatch in the statistical model
used for analysis. This mismatch can potentially inflate the reliability
of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias
into the trait's estimates \citep{Ackerman_1989}. Similarly, failing to
account for hierarchical (grouping) structures reduces the precision of
model parameter estimates, further amplifying the overestimation of
reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
also undermine the trait's reliability and ultimately compromise the
validity of the trait's estimates \citep[pp.~2]{Perron_et_al_2015}.

Fortunately, the same literature offers solutions for addressing these
issues. \citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend
integrating judges' biases into the BTL model. Moreover, the literature
advocates for the incorporation of relevant hierarchical structures into
the statistical model to account for these dependencies. Together, these
additions can result in a model resembling a Multilevel Structural
Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
multidimensional or two-parameter logistic Item Response Theory (IRT)
model \citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
and statistical treatment of judges' biases.

@@

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and IRT
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
recommend conducting these analyses and tests within a structural model.
A structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the BTL scores
and their uncertainties simultaneously, rather than treating them as
separate elements.

>>>>

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. The present study primarily focuses on two: the
overreliance on the assumptions of Thurstone's Case V in the statistical
analysis of CJ data, and the apparent disconnect between CJ's approach
to trait measurement and hypothesis testing. The following sections
begin with a brief overview of Thurstone's theory and a detailed
examination of these issues. Subsequently, the study introduces a
theoretical model for CJ that builds upon Thurstone's theory, alongside
its statistical translation, designed to address the two concerns
simultaneously.

@@

In its most general form, Thurstone's theory addresses pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their perception of the stimulus
trait. According to the theory, the discriminal process for each
stimulus follows a Normal distribution, where its mode (mean), called
the \emph{modal discriminal process}, indicates the stimulus position on
the trait continuum, while its dispersion, known as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates the discriminal
processes along a quality trait continuum for two written texts. The
figure shows that these processes individually follow a Normal
distribution. Furthermore, it shows that the modal discriminal process
for Text B is positioned further along the continuum than Text A
\((S_{B} > S_{A})\). Additionally, the figure highlights the broader
distribution of Text B compared to Text A as a result of its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

\caption{\label{fig-discriminal_process}Discriminal processes for two
written texts}

@@

However, since the discriminal processes of the stimuli are not directly
observable, the theory introduces the \emph{law of comparative
judgment}. This law posits that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This emphasizes that the relative
distance between stimuli, rather than their absolute positions on the
continuum, determines the outcome of the pairwise comparison. Indeed,
the theory assumes that the observed dichotomous outcome arises from the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}. Since the individual
discriminal processes follow a Normal distribution, the discriminal
difference also follows a Normal distribution \citep{Andrich_1978}, with
its mode (mean) representing the relative separation between the
stimuli, while its dispersion reflects the variability in that relative
separation.

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}. The figure illustrates that the
judge perceives Text B as having significantly higher quality than Text
A, as indicated by their discriminal difference \((S_{B} - S_{A} > 0)\)
and the shaded gray area under the curve \(P(B > A)\). Therefore, the
dichotomous outcome of this comparison would likely favor Text B over
Text A.

@@

\caption{\label{fig-discriminal_difference}Discriminal difference for
two written texts}

@@

\section{The two critical issues in CJ
literature}\label{sec-theory-issues}

This section examines the two critical issues in the CJ literature that
serve as the primary focus of this study. The first is the overreliance
on Thurstone's Case V assumptions in the statistical analysis of CJ
data. The second is the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing.

@@

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. Moreover, he warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions might be problematic when complex traits or
heterogeneous stimuli are involved \citep[pp.~376]{Thurstone_1927a}.
Consequently, given that modern CJ applications frequently involve such
traits and stimuli, two main assumptions of Case V and, by extension, of
the BTL model may not consistently hold in theory or practice: the
assumption of equal dispersion and zero correlation between stimuli.

\subsubsection{The assumption of equal dispersions between
stimuli}\label{the-assumption-of-equal-dispersions-between-stimuli}

The discriminal dispersions of stimuli are crucial in determining the
dichotomous outcomes of pairwise comparisons. Discrepancies in these
dispersions shape the distribution of the discriminal difference,
directly influencing the comparison outcome. For instance,
Figure~\ref{fig-dispersion} illustrates how more uncertainty in the
trait perception of a text over another broadens the distribution of
their discriminal difference. This broadening affects the area under the
curve, \(P(B>A)\), determining the comparison outcome. Additionally, the
figure shows that when there are no discrepancies in the texts'
discriminal dispersions, meaning the \emph{texts have equal dispersions}
\((\sigma_{B}-\sigma_{A}=0)\), the likelihood that the discriminal
difference distinctly favors one text over another increases.

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/dispersion.png}

\caption{\label{fig-dispersion}The effect of different dispersions
between stimuli on the distribution of the discriminal difference}

However, Thurstone contended that this assumption may not hold when
researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
introduce judgment discrepancies due to their unique features. Indeed,
evidence of such violation may already exist in the CJ literature as
misfit statistics. These statistics measure the judgment discrepancies
associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than
comparisons involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that the discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
their discriminal processes also exhibit more variation than other
texts. Notably, this reasoning also applies to ``misfit'' judges, whose
evaluations reflect substantial deviations from the shared consensus due
to the unique characteristics of the stimuli or the judges themselves.

\subsubsection{The assumption of zero correlation between
stimuli}\label{the-assumption-of-zero-correlation-between-stimuli}

Similar to the discriminal dispersions, the correlation between the
discriminal processes is also crucial in determining the dichotomous
outcomes of pairwise comparisons. This correlation shapes the
distribution of the discriminal difference, directly influencing the
comparison outcome. For instance, Figure~\ref{fig-correlation}
illustrates how the stronger dependence of the trait perception in one
text over another narrows the distribution of their discriminal
difference. This narrowing impacts the area under the curve, \(P(B>A)\),
determining the comparison outcome. Additionally, the figure shows that
when the trait perception in one text is independent of another, meaning
the \emph{texts have zero correlation}, the likelihood that the
discriminal difference distinctly favors one text over another is lower
compared to when the texts are highly correlated.

\caption{\label{fig-correlation}The effect of correlation on the
distribution of the discriminal difference}

Notably, Thurstone's Case V and the BTL model assume independent trait
perceptions across stimuli. Thurstone attributed this independence to
the cancellation of potential judges' biases, driven by two opposing and
equally weighted effects occurring during the pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
demonstrated this cancellation using the BTL model under the assumption
of discriminal processes with additive biases. However, it is easy to
imagine at least two scenarios where the zero correlation assumption
almost certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

@@

@@
